\cursedforest\ - A Random Forest Implementation for ``Big'' and ``Wide'' Data \\ Supplementary Information

### 2017-01-12 doc/publications/plos16/plos16.tex
Update my section a bit

% Use Unicode characters when possible
\usepackage[utf8x]{inputenc}

@@

\newcommand{\cursedforest}{\textsc{CursedForest}\xspace}
\newcommand{\ranger}{\textsc{Ranger}\xspace}
\newcommand{\mtry}{\texttt{mtry}\xspace}
\newcommand{\ntree}{\texttt{ntree}\xspace}

@@

%% Additional packages
\usepackage{footmisc}
\usepackage{multirow}

@@

% Please keep the abstract below 300 words
\section{Abstract}
We introduce \cursedforest, an implementation of random forests designed to handle data that may have an extremely large
number of variables. We discuss the utility of random forests for machine learning problems. We show on simulated and
real data that the novel implementation can achieve reasonable prediction accuracy and variable importance estimates
even in setting where other implementations, including some designed for large data, fail.

@@

The most popular random forest software packages are R-based, although current implementations do not scale well with
increasing feature size~\cite{Wright.and.Ziegle.2016}.  This is despite R now supporting long vectors (length greater
than $2^{31}-1$), and the \textsc{randomForest} implementation in R lending itself to parallelization to grow multiple
trees simultaneously~\cite{Liaw.and.Weiner.2002} and them combine them.  More suitable implementations for ``wide''
genomic data have been developed in other programming languages. \textsc{CloudForest} \cite{Bressler2015} written in Go,
achieves fast run times by effective use of the CPU cache, optimizing for different classes of features and efficiently
multi-threading.  Similarly, \ranger~\cite{Wright.and.Ziegle.2016} written in C++ with an R front-end is
specifically optimized for dealing with large data sets.

@@

As is standard with Spark applications, we store our data in a Resilient Distributed Dataset (RDD), where an RDD is
essentially a collection of elements. In the case of Spark ML, each element in the RDD is a sample. RDDs contribute to
the scalability of Spark as they can distributed across multiple nodes and operated on in parallel. Even as we add more
samples to a dataset, Spark can still simply schedule extra tasks to handle the additional items in the RDD.

@@

So in the case of adding more samples, Spark ML can simply create more tasks, keeping memory consumption within the
cluster's bounds. However, as the dimensionality of each sample grows, the memory requirements of the job increases to
enable these increasingly large vectors to be loaded into memory.

On the other hand, \cursedforest is specifically designed to handle wide ``cursed'' data. It avoids the relation between
memory and dimensionality by avoiding calculations that rely on entire feature vectors and taking the parallelization
work down to the level of the individual features.  For example, for each node of a tree, \cursedforest will distribute
tasks that consist of single features (variants), for every individual.  Each of these tasks will calculate the
information gain for that specific feature.  Once these tasks have completed, the results are reduced to return the
feature which gives the greatest information gain.  This process is then repeated until \cursedforest has created the
entire decision tree.

The current implementation of \cursedforest uses an ``Gini impurity'' criteria for splitting. Let $f_q$ be the fraction
of items labelled with value $q$ where $q = 1, \ldots, Q$ at a node. The Gini impurity is
\[
I_G(f) = \Sigma_{q = 1}^Q f_q ( 1 − f_q ), 
\]
which is a minimum when all observatkions at the node are in the same class.

@@

\subsubsection{The Cancer Genome Atlas}
We obtained The Cancer Genome Atlas (TCGA) BRCA data as VCF files from the GDC data portal.  This data is in the
category ``Simple Nucleotide Variation''.  After downloading, we use a custom script to name the samples in the files
and filter for only the "Normal" tissue column.  We then use GATK to merge the files into a single VCF file.

@@

\subsection{Parameter settings}
We consider the parameter setting for the random forest algorithm. We use the R notation from the \textsc{randomForest}
package \cite{Liaw.and.Weiner.2002} which incorporates the original Fortran code by Brieman and Cutler.  We incorporate
the advice of \cite{Liaw.and.Weiner.2002}, which we have found mirrors our own experience:
\begin{itemize}
\item \ntree{}  -- the number of trees.  The number of trees necessary for good performance grows with the number of
  predictors.  \cite{Liaw.and.Weiner.2002} suggest that a high \ntree is necessary to get stable estimates of variable
  importance and proximity; however, even though the variable importance measures may vary from run to run, we note that
  it is possible for a random forest model to have a poorer fit and still have an accurate ranking of variable
  importance;
\item \mtry{}  -- the number of variables considered at each split (if \mtry=$p$, we have a boosted decision
  tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger \mtry may give
  better performance;
\item the size and complexity of the individual trees is controlled in
  \textsc{randomForest} by setting \texttt{nodesize}, the minimum size of
  terminal nodes. It is controlled in Spark ML by setting
  \texttt{maxDepth}, the maximum depth of each tree in the
  forest\footnote{The Spark ML documentation \cite[]{Spark.2016} sets
    \texttt{maxDepth} to 4 in their classification example. This may
    give an estimate with a higher bias and, as such, may be a poor
    choice. See the discussion in \cite{Dietterich.2002} where it is
    shown that bagging may reduce bias as well as variance but gives a
    better result for low bias learners. \cite{Dietterich.2002} notes
    that if the boostrap replicate approximation were correct, then
    bagging would reduce variance without changing bias.} In the TCGA
  example the RandomForests code produced trees with depths of 30+.
\end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 2: Feature Selection on real data 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest\ selects previously demonstrated cancer variants}
We consider the question of feature discovery on the Cancer Genome Atlas (TCGA) BRCA dataset.  We build this model on
VCF files retrieved from the TCGA data portal.  For the label, we use the cell type, tumor or normal.

>>>>

% Use Unicode characters when possible
\usepackage[utf8]{inputenc}

@@

\newcommand{\cursedforest}{\textsc{CursedForest}\xspace}
\newcommand{\mtry}{\texttt{mtry}\xspace}
\newcommand{\ntree}{\texttt{ntree}\xspace}

@@

%% Additional packages
\usepackage{footmisc}
\usepackage{multirow}
\usepackage[T1]{fontenc}

@@

% Please keep the abstract below 300 words
\section{Abstract}
The abstract.

@@

The most popular random forest software packages are R-based, although current implementations do not scale well with
increasing feature size~\cite{Wright.and.Ziegle.2016}.  This is despite R now supporting long vectors (length greater
than $2^{31}-1$), and the \textsc{randomForest} implementation in R lending itself to parallelization to grow multiple
trees simultaneously~\cite{Liaw.and.Weiner.2002} and them combine them.  More suitable implementations for ``wide''
genomic data have been developed in other programming languages. \textsc{CloudForest} \cite{Bressler2015} written in Go,
achieves fast run times by effective use of the CPU cache, optimizing for different classes of features and efficiently
multi-threading.  Similarly, \textsc{Ranger}~\cite{Wright.and.Ziegle.2016} written in C++ with an R front-end is
specifically optimized for dealing with large data sets.

@@

As is standard with Spark applications, we store our data in a Resilient Distributed Dataset (RDD), where an RDD is
essentially a collection of elements. In the case of Spark ML, each element in our RDD is a sample. RDDs contribute to
the scalability of Spark as they can distributed across multiple nodes and operated on in parallel. Even as we add more
samples to a dataset, Spark can still simply schedule extra tasks to handle the additional items in the RDD.

@@

So in the case of adding more samples, Spark ML can simply create more tasks, keeping memory consumption within the
cluster's bounds. However, as the dimensionality of each sample grows, the memory requirements of the job increases to
enable these increasingly large vectors to be loaded into memory.
\marginpar{not clear. Are RDDs the problem or the solutuion Aidan: Better?}

On the other hand, \cursedforest\ is specifically designed to handle wide ``cursed'' data. It avoids the relation
between memory and dimensionality by avoiding calculations that rely on entire feature vectors.

Compared to Spark ML, \cursedforest\ further parallelizes work down to individual features.  For example, for each node
of a tree, \cursedforest\ will distribute tasks that consist of single features (variants), for every individual.  Each
of these tasks will calculate the information gain for that specific feature.  Once these tasks have completed, the
results are reduced to return the feature which gives the greatest information gain.  This process is then repeated
until \cursedforest\ has created the entire decision tree.

The current implementation of \cursedforest\ uses an ``Gini impurity''
criteria for splitting. Let $f_q$ be the fraction of items labelled
with value $q$ where $q = 1, \ldots, Q$ at a node. The Gini impurity is
\[
I_G(f) = \Sigma_{q = 1}^Q f_q ( 1 - f_q ), 
\]
which is a minimum when all observatkions at the node are in the same class.

@@

\subsubsection{Bone mineral density dataset}
We obtained a VCF file from the authors of \cite{Duncan2011}, which contains variant data from 1,936 postmenopausal women.
This dataset covers approximately 7 million variant positions. We have information about the bone mineral density (BMD)
for each individual, where a low BMD is a major predisposing factor to fracture \cite{Duncan2011}.

@@

\subsection{Parameter settings}
We consider the parameter setting for the random forest algorithm. We use the R notation from the \textsc{randomForest}
package \cite{Liaw.and.Weiner.2002} which incorporates the original Fortran code by Brieman and Cutler.  We incorporate
the advice of \cite{Liaw.and.Weiner.2002}, which we have found mirrors our own experience:
\begin{itemize}
\item \ntree{}  -- the number of trees.  The number of trees necessary for good performance grows with the number of
  predictors.  \cite{Liaw.and.Weiner.2002} suggest that a high \ntree\ is necessary to get stable estimates of variable
  importance and proximity; however, even though the variable importance measures may vary from run to run, we note that
  it is possible for a random forest model to have a poorer fit and still have an accurate ranking of variable
  importance;
\item \mtry{}  -- the number of variables considered at each split (if \mtry=$p$, we have a boosted decision
  tree model).  If one has a very large number of variables but expects only very few to be ``important'', using larger \mtry\ may give
  better performance;
\item the size and complexity of the individual trees is controlled in
  \textsc{randomForest} by setting \texttt{nodesize}, the minimum size of
  terminal nodes. It is controlled in Spark ML by setting
  \texttt{maxDepth}, the maximum depth of each tree in the
  forest\footnote{The Spark ML documentation \cite[]{Spark.2016} sets
    \texttt{maxDepth} to 4 in their classification example. This may
    give an estimate with a higher bias and, as such, may be a poor
    choice. See the discussion in \cite{Dietterich.2002} where it is
    shown that bagging may reduce bias as well as variance but gives a
    better result for low bias learners. \cite{Dietterich.2002} notes
    that if the boostrap replicate approximation were correct, then
    bagging would reduce variance without changing bias.} In the TCGA
  example the RandomForests code produced trees with depths of 30+.
\end{itemize}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 2: Feature Selection on real data 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest\ selects previously demonstrated bone-density variants}
We consider the problem of feature discovery on a genomic dataset for a specific phenotype.
In this case we investigate the ability of  ability to locate variants in the same loci as genes previously demonstrated to
be associated with a high or low bone mass density (BMD) and rank these variants as being "important".
From a random forest model built on this data, we produce a list of the to 5,000 features. 
Applying the Mann–Whitney U test to these fetures we observe a p-value of x, supporting our
hypothesis that features in close proximity to the known BMD genes are Contributing more information to our classification model.

One observation we make is that of the most-important features---for example, the top ten variants ranked by Gini impurity---is
that not all of these are variants are in close proximity of known BMD genes.
One of these is COLEC10. Featuring a transcription factor binding site for FOXL1, COLEC10 is also reported as containing a SNP
associated with BMD \cite{Liu2008}.
